A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-08-26 masters_thesis/si_paper/paper.tex
More of the minor revisions

A general rule of thumb for when this architecture is useful is when the
combined function evaluation time is greater than the communication overhead
for a single iteration; i.e., $f*p > o+f$, where $f$ is the function evaluation
time, $p$ is the desired number of particles (in our case $p$ is equal to the
number of available processors), and $o$ is the amount of overhead per
iteration.  The amount of overhead per iteration will vary based on the
communication topology and the number of particles, so this is only an
approximation, but it can provide rough estimates of function evaluation times
at which this parallel architecture should be used.

@@

Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
10\right), z=x-c\]

Sphere is initialized in $[-50,50]^D$ and is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]

Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]

Griewank is initialized in $[-600,600]^D$ and is defined as:
\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]

And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
x_{i+1})+.7)\]

In computing the branch statistics in \sec{statistics}, we also used the
Rosenbrock function.  That function is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D \left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]

@@

We frequently present tables summarizing our results.  In each table, we bold
the ``best'' method, meaning that it has at least a 90\% success rate and its
average time to completion is faster than all other methods that have at least
a 90\% success rate.  We used a double-sided t-test to test for statistical
significance in these results, and all bolded values are statistically
significant with a p-value of less than $10^{-5}$.  In some instances two
values are bolded because they are statistically tied, while both are
significantly better than all other methods.

@@

Schwefel 2.21 is a function similar to Sphere, but Schwefel 2.21 is benefited
more by larger swarms than Sphere is.  Thus our speculative algorithms often
fail to outperform Standard with 240 processors because the simple speculative
methods only have 30 particles.  As we will show later, when we use 800
processors at 50 dimensions, our methods perform much better.  However, the
pruned versions of our methods have 80 particles and thus are able to
outperform Standard even with only 240 processors.  Because the graph of
Schwefel 2.21 looks very similar to that of Sphere, we simply present all of
our results in \tabref{schwefel-20}.

@@

We have given five different possible approaches to speculative evaluation,
each of which has different properties.  These approaches perform differently
on different functions and at different swarm sizes, as would be expected by
the No Free Lunch Theorem for
Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
evaluation of the premature convergence properties of these methods on
deceptive functions when a smaller number of processors is available.  We have
also shown evidence that when many processors are available by far the best
thing to do in most cases is to speculate as far ahead as the extra processors
allow.  This point is shown in a single graph in \fig{griewank-all}.  In that
case, the extra processors in standard parallelizations provided \emph{no
appreciable benefit.}  In contrast, speculative methods allow additional
processors to provide arbitrarily high speed ups, as the more processors are
available the farther ahead the algorithm can speculate.

>>>>

A general rule of thumb for when this architecture is useful is when the
combined function evaluation time is greater than the communication overhead
for a single iteration; i.e., $t*p > o+t$, where $t$ is the function evaluation
time, $p$ is the desired number of particles (in our case $p$ is equal to the
number of available processors), and $o$ is the amount of overhead per
iteration.  The amount of overhead per iteration will vary based on the
communication topology and the number of particles, so this is only an
approximation, but it can provide rough estimates of function evaluation times
at which this parallel architecture should be used.

@@

Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
\[f_{Rastrigin}(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi
z_i\right) + 10\right), z=x-c\]

Sphere is initialized in $[-50,50]^D$ and is defined as:
\[f_{Sphere}(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]

Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
\[f_{Schwefel}(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]

Griewank is initialized in $[-600,600]^D$ and is defined as:
\[f_{Griewank}(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]

And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
\[f_{Bohachevsky}(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi
x_i)-.4\cos(4\pi x_{i+1})+.7)\]

In computing the branch statistics in \sec{statistics}, we also used the
Rosenbrock function.  That function is defined as:
\[f_{Rosenbrock}(\Vec{x}) = \sum_{i=1}^D
\left(100(x_{i+1}-x_i^2)+(x_i-1)^2\right)\]

@@

We frequently present tables summarizing our results.  In each table, we bold
the ``best'' method, meaning that it has at least a 90\% success rate and its
average time to completion is faster than all other methods that have at least
a 90\% success rate\footnote{``Success rate'' here means that the algorithm
reached some predefined value which is mentioned in the table or figure
wherever a success rate is reported.}.  We used a double-sided t-test to test
for statistical significance in these results, and all bolded values are
statistically significant with a p-value of less than $10^{-5}$.  In some
instances two values are bolded because they are statistically tied, while both
are significantly better than all other methods.

@@

Schwefel 2.21 is a function similar to Sphere (in that they are both unimodal),
but Schwefel 2.21 is benefited more by larger swarms than Sphere is (as it is
harder to optimize).  Thus our speculative algorithms often fail to outperform
Standard with 240 processors because the simple speculative methods only have
30 particles.  As we will show later, when we use 800 processors at 50
dimensions, our methods perform much better.  However, the pruned versions of
our methods have 80 particles and thus are able to outperform Standard even
with only 240 processors.  Because the graph of Schwefel 2.21 looks very
similar to that of Sphere, we simply present all of our results in
\tabref{schwefel-20}.

@@

We have given five different possible approaches to speculative evaluation,
each of which has different properties.  These approaches perform differently
on different functions and at different swarm sizes, as would be expected by
the No Free Lunch Theorem for
Optimization~\citep{wolpert-1997-nfl-for-optimization}.  We have given a brief
evaluation of the premature convergence properties of these methods on
deceptive functions when a smaller number of processors is available.  We have
also shown evidence that when many processors are available by far the best
thing to do in most cases is to speculate as far ahead as the extra processors
allow.  This point is shown in a single graph in \fig{griewank-all}.  In that
case, the extra processors in standard parallelizations provided \emph{no
appreciable benefit.}  In contrast, speculative methods allow additional
processors to provide additional speed ups, as the more processors are
available the farther ahead the algorithm can speculate.
